\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-2\baselineskip}

### 2017-06-08 paper/ms.tex
words and edits about factorization and solving

\subsection{Cholesky factorization}

@@

\subsection{Scalable factorization of the celerite model}

The \celeriteterm\ model described in \sect{celerite} can be exactly
represented as a rank $R = 2\,J$ semiseparable matrix\footnote{We note that if
some \celeriteterm\ terms are real ($b_j=0$ and $d_j=0$) then these terms only
add one to the rank.
If a \celeriteterm\ model has $J_\mathrm{real}$ real and $J_\mathrm{complex}$
general terms, the semiseparable rank is only $J_\mathrm{real} +
2\,J_\mathrm{complex}$.}
where the components of the relevant matrices are
\begin{eqnarray}
U_{n,2\,j-1} &=& a_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) +
    b_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) \\
U_{n,2\,j} &=& a_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) -
    b_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) \\
V_{m,2\,j-1} &=& e^{c_j\,t_m}\,\cos(d_j\,t_m) \\
V_{m,2\,j} &=& e^{c_j\,t_m}\,\sin(d_j\,t_m) \\
A_{n,n} &=& \sigma_n^2 + \sum_{j=1}^J a_j \quad.
\end{eqnarray}
Na\"ive implementation of the algorithm in \eq{algorithm} to this system will
result in numerical overflow and underflow issues in many realistic cases
because both $U$ and $V$ have factors of the form $e^{\pm c_j t}$, where $t$
can be any arbitrarily large real number.
This issue was also noted by \citet{Ambikasaran:2015} and it can be avoided
with a reparameterization similar to the one they use.
We define the following pre-conditioned variables
\begin{eqnarray}
\tilde{U}_{n,2\,j-1} &=& a_j\,\cos(d_j\,t_n) + b_j\,\sin(d_j\,t_n) \\
\tilde{U}_{n,2\,j} &=& a_j\,\sin(d_j\,t_n) - b_j\,\cos(d_j\,t_n) \\
\tilde{V}_{m,2\,j-1} &=& \cos(d_j\,t_m) \\
\tilde{V}_{m,2\,j} &=& \sin(d_j\,t_m) \\
\tilde{W}_{n,2\,j-1} &=& e^{-c_j\,t_n}\,W_{n,2\,j-1} \\
\tilde{W}_{n,2\,j} &=& e^{-c_j\,t_n}\,W_{n,2\,j}
\end{eqnarray}
and the set of variables
\begin{eqnarray}
\phi_{n,2\,j-1} = \phi_{n,2\,j} = e^{-c_j\,(t_n - t_{n-1})}
\end{eqnarray}
with the constraint
\begin{eqnarray}
\phi_{1,2\,j-1} = \phi_{1,2\,j} = 0 \quad.
\end{eqnarray}
Using this parameterization, we find the following, numerically stable,
algorithm to compute the Cholesky factorization of a \celeriteterm\ model
\begin{eqnarray}
    S_{n,j,k} &=& \phi_{n,j}\,\phi_{n,k}\,\left[S_{n-1,j,k} +
    D_{n-1,n-1}\,\tilde{W}_{n-1,j}\,\tilde{W}_{n-1,k}\right] \nonumber\\
D_{n,n} &=& A_{n,n} -
    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} \tilde{U}_{n,j}\,S_{n,j,k}\,\tilde{U}_{n,k}
    \nonumber\\
\tilde{W}_{n,j} &=& \frac{1}{D_{n,n}}\left[ \tilde{V}_{n,j} -
    \sum_{k=1}^{2\,J}\tilde{U}_{n,k}\,S_{n,j,k} \right] \quad.
\end{eqnarray}
As before, a single iteration of this algorithm requires $\mathcal{O}(J^2)$
operations and it is repeated $N$ times for a total scaling of
$\mathcal{O}(N\,J^2)$.
Furthermore, $\tilde{V}_{n,j}$ and $A_{n,n}$ can be updated in place and used
to represent $\tilde{W}_{n,j}$ and $D_{n,n}$, giving a total memory footprint
of $(6\,J + 1)\,N + J\,(J-1)/2$ floating point numbers.

\subsection{Solving}

After computing the Cholesky factorization of the matrix $K$, we can apply
the inverse of $K$ and compute its log-determinant in $\mathcal{O}(N\,J)$ and
$\mathcal{O}(N)$ operations respectively.
The log-determinant of $K$ is given by
\begin{eqnarray}
    \ln \det K = \sum_{n=1}^N \ln D_{n,n}
\end{eqnarray}
where, if $K$ is positive semi-definite, $D_{n,n}>0$ for all $n$.

The inverse of $K$ can be applied using the relationship
\begin{eqnarray}
\bvec{z} = K^{-1}\,\bvec{y} = {(L^\T)}^{-1}\,D^{-1}\,L^{-1}\,\bvec{y} \quad.
\end{eqnarray}
First, we solve for $\bvec{z}^\prime = L^{-1}\,\bvec{y}$ using the following forward
substitution algorithm
\begin{eqnarray}
    f_{n,j} &=& \phi_{n,j}\,\left[f_{n-1,j} +
    \tilde{W}_{n-1,j}\,z_{n-1}^\prime\right] \nonumber\\
    z_n^\prime &=& y_n - \sum_{j=1}^{2\,J} \tilde{U}_{n,j}\,f_{n,j}
\end{eqnarray}
where $f_{0,j} = 0$ for all $j$.
Using this result and the following backward substitution algorithm, we can
compute $\bvec{z}$
\begin{eqnarray}
g_{n,j} &=& \phi_{n+1,j}\,\left[g_{n+1,j} +
    \tilde{U}_{n+1,j}\,z_{n+1}\right] \nonumber\\
z_n &=& \frac{z_n^\prime}{D_{n,n}} - \sum_{j=1}^{2\,J} \tilde{W}_{n,j}\,g_{n,j}
\end{eqnarray}
where $g_{N+1,j} = 0$ for all $j$.
The total cost of the forward and backward substitution passes scales as
$\mathcal{O}(N\,J)$.

>>>>

\subsection{Cholesky factorization of PSD semiseparable matrices}

@@

After computing the Cholesky factorization of this matrix $K$, we can also
apply its inverse and compute its log-determinant in $\mathcal{O}(N\,R)$ and
$\mathcal{O}(N)$ operations respectively.
The log-determinant of $K$ is given by
\begin{eqnarray}
    \ln \det K = \sum_{n=1}^N \ln D_{n,n}
\end{eqnarray}
where, since $K$ is PSD, $D_{n,n}>0$ for all $n$.

The inverse of $K$ can be applied using the relationship
\begin{eqnarray}
\bvec{z} = K^{-1}\,\bvec{y} = {(L^\T)}^{-1}\,D^{-1}\,L^{-1}\,\bvec{y} \quad.
\end{eqnarray}
First, we solve for $\bvec{z}^\prime = L^{-1}\,\bvec{y}$ using forward
substitution
\begin{eqnarray}\eqlabel{fwd}
    f_{n,j} &=& f_{n-1,j} + W_{n-1,j}\,z_{n-1}^\prime \nonumber\\
    z_n^\prime &=& y_n - \sum_{j=1}^{R} U_{n,j}\,f_{n,j}
\end{eqnarray}
where $f_{0,j} = 0$ for all $j$.
Then, using this result and backward substitution, we compute $\bvec{z}$
\begin{eqnarray}\eqlabel{bwd}
g_{n,j} &=& g_{n+1,j} + U_{n+1,j}\,z_{n+1} \nonumber\\
z_n &=& \frac{z_n^\prime}{D_{n,n}} - \sum_{j=1}^{R} W_{n,j}\,g_{n,j}
\end{eqnarray}
where $g_{N+1,j} = 0$ for all $j$.
The total cost of the forward and backward substitution passes scales as
$\mathcal{O}(N\,R)$.

The algorithm derived here is a general method for factorizing, solving, and
computing the determinant of PSD rank-$R$ semiseparable matrices.
As we will show below, with some care, this method can be used to compute
\eq{gp-likelihood} for \celeriteterm\ models.
In practice, when $K$ is PSD and this method is applicable, we find that this
method is about a factor of 20 faster than an optimized implementation of the
earlier, more general method from \citet{Ambikasaran:2015}.

\subsection{Scalable computation of celerite models}

The \celeriteterm\ model described in \sect{celerite} can be exactly
represented as a rank $R = 2\,J$ semiseparable matrix\footnote{We note that if
some \celeriteterm\ terms are real ($b_j=0$ and $d_j=0$) then these terms only
add one to the rank.
Therefore, if a \celeriteterm\ model has $J_\mathrm{real}$ real and
$J_\mathrm{complex}$ general terms, the semiseparable rank is only
$J_\mathrm{real} + 2\,J_\mathrm{complex}$.}
where the components of the relevant matrices are
\begin{eqnarray}\eqlabel{system}
U_{n,2\,j-1} &=& a_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) +
    b_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) \nonumber\\
U_{n,2\,j} &=& a_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) -
    b_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) \nonumber\\
V_{m,2\,j-1} &=& e^{c_j\,t_m}\,\cos(d_j\,t_m) \nonumber\\
V_{m,2\,j} &=& e^{c_j\,t_m}\,\sin(d_j\,t_m) \nonumber\\
A_{n,n} &=& \sigma_n^2 + \sum_{j=1}^J a_j \quad.
\end{eqnarray}
A na\"ive implementation of the algorithm in \eq{algorithm} for this system
will result in numerical overflow and underflow issues in many realistic cases
because both $U$ and $V$ have factors of the form $e^{\pm c_j t}$, where $t$
can be any arbitrarily large real number.
This issue can be avoided by reparameterizing the semiseparable representation
from \eq{system}.
%A similar issue was also noted by \citet{Ambikasaran:2015} and it can be
%avoided using a reparameterization.
We define the following pre-conditioned variables
\begin{eqnarray}\eqlabel{reparam}
\tilde{U}_{n,2\,j-1} &=& a_j\,\cos(d_j\,t_n) + b_j\,\sin(d_j\,t_n) \nonumber\\
\tilde{U}_{n,2\,j} &=& a_j\,\sin(d_j\,t_n) - b_j\,\cos(d_j\,t_n) \nonumber\\
\tilde{V}_{m,2\,j-1} &=& \cos(d_j\,t_m) \nonumber\\
\tilde{V}_{m,2\,j} &=& \sin(d_j\,t_m) \nonumber\\
\tilde{W}_{n,2\,j-1} &=& e^{-c_j\,t_n}\,W_{n,2\,j-1} \nonumber\\
\tilde{W}_{n,2\,j} &=& e^{-c_j\,t_n}\,W_{n,2\,j}
\end{eqnarray}
and the new set of variables
\begin{eqnarray}
\phi_{n,2\,j-1} = \phi_{n,2\,j} = e^{-c_j\,(t_n - t_{n-1})}
\end{eqnarray}
with the constraint
\begin{eqnarray}
\phi_{1,2\,j-1} = \phi_{1,2\,j} = 0 \quad.
\end{eqnarray}
Using this parameterization, we find the following, numerically stable,
algorithm to compute the Cholesky factorization of a \celeriteterm\ model
\begin{eqnarray}
    S_{n,j,k} &=& \phi_{n,j}\,\phi_{n,k}\,\left[S_{n-1,j,k} +
    D_{n-1,n-1}\,\tilde{W}_{n-1,j}\,\tilde{W}_{n-1,k}\right] \nonumber\\
D_{n,n} &=& A_{n,n} -
    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} \tilde{U}_{n,j}\,S_{n,j,k}\,\tilde{U}_{n,k}
    \nonumber\\
\tilde{W}_{n,j} &=& \frac{1}{D_{n,n}}\left[ \tilde{V}_{n,j} -
    \sum_{k=1}^{2\,J}\tilde{U}_{n,k}\,S_{n,j,k} \right] \quad.
\end{eqnarray}
As before, a single iteration of this algorithm requires $\mathcal{O}(J^2)$
operations and it is repeated $N$ times for a total scaling of
$\mathcal{O}(N\,J^2)$.
Furthermore, $\tilde{V}_{n,j}$ and $A_{n,n}$ can be updated in place and used
to represent $\tilde{W}_{n,j}$ and $D_{n,n}$, giving a total memory footprint
of $(6\,J + 1)\,N + J\,(J-1)/2$ floating point numbers.

The log-determinant of $K$ can still be computed as in the previous section,
but we must update the forward and backward substitution algorithms to account
for the reparameterization in \eq{reparam}.
The forward substitution with the reparameterized variables is
\begin{eqnarray}
    f_{n,j} &=& \phi_{n,j}\,\left[f_{n-1,j} +
    \tilde{W}_{n-1,j}\,z_{n-1}^\prime\right] \nonumber\\
    z_n^\prime &=& y_n - \sum_{j=1}^{2\,J} \tilde{U}_{n,j}\,f_{n,j}
\end{eqnarray}
where $f_{0,j} = 0$ for all $j$, and the backward substitution algorithm
becomes
\begin{eqnarray}
g_{n,j} &=& \phi_{n+1,j}\,\left[g_{n+1,j} +
    \tilde{U}_{n+1,j}\,z_{n+1}\right] \nonumber\\
z_n &=& \frac{z_n^\prime}{D_{n,n}} - \sum_{j=1}^{2\,J} \tilde{W}_{n,j}\,g_{n,j}
\end{eqnarray}
where $g_{N+1,j} = 0$ for all $j$.
The only differences, when compared to Equations~(\eqref{fwd})
and~(\eqref{bwd}) is the multiplication by $\phi_{n,j}$ in the recursion
relations for $f_{n,j}$ and $g_{n,j}$.
As before, the computational cost of this solve scales as $\mathcal{O}(N\,J)$.
